\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-18 paper/algorithm.tex
paper: algorithm helpers clean up, add text about planner

\todo{formatting of plan cache semantics needs to be fixed}
\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
                                           P = \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
                                                                   \end{aligned}
                                           \right\}
                                         \end{aligned}
    % \in 
  \end{align*}
  \caption{Semantics of the plan cache.}
\end{figure}

\todobox{reorder these and explain algorithm, which right now is cutoff}{
The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache

@@

Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
}
\begin{algorithm}
\todo{fix input descriptions and small error in optjoin}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

>>>>

\begin{figure}
\todo{formatting of plan cache semantics needs to be fixed}
\todo{I think the cache update step is not quite right, since p seems to be a single plan, but it is really a set}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
                                           P = \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
                                                                   \end{aligned}
                                           \right\}
                                         \end{aligned}
    % \in 
  \end{align*}
  \caption{Semantics of the plan cache.}
\end{figure}

@@

Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top-level planner, shown in Algorithm~\ref{algo:top-level-planner}.

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache. If no selections are available, a 
simple scan is added to the plan cache. Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced. This set is now extended for any necessary grouping and aggregation
operations. After a similar step has been taken for projections, the planner returns the cheapest plan from the cache.

\begin{algorithm}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

### 2017-02-18 paper/algorithms/plan-helpers.tex
paper: algorithm helpers clean up, add text about planner

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan. $C_l$ is a set of attributes that must be imputed in the output of this query plan. $C_g$ is the set of attributes which are used in the final plan.}
    \Ensure{A set of query plans $Q$ with added imputation.}
    \Function{AddImpute}{$q, C_l, C_g$}
    \State $D_{must} \gets \Call{Dirty}{q} \cap C_l$
    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_g)$
    \State $Q \gets \emptyset$
    \If{$D_{must} = \emptyset$}
    \State $Q \gets Q \cup \{q\}$
    \Else
    \State $Q \gets Q \cup \{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}$
    \EndIf

@@

    \Require{$t$ is a table and $\phi$ is a filter predicate.}
    \Ensure{A set of plans for filtering $t$ with $\phi$.}
    \Function{OptFilter}{$t, \phi, C_g$}
    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_g}\}$
    \EndFunction

@@

    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, and $C_g$ is a set of visible attributes.}
    \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
    \Function{OptGroupBy}{$q, G, A, C_g$}
    \State $C \gets G \cup \Call{Attrs}{A}$
    \State $Q' \gets \Call{AddImpute}{q, C, C_g}$
    \State \Return $\{\Call{GroupBy}{q', G, A} ~|~ q' \in Q'\}$
    \EndFunction

@@

    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates.}
    \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
    \Function{OptJoin}{$Q, T, J, C_g$}
    \For{$size \in 2...|T|$}
    \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
    \For{$t \in S$}
    \State $S' \gets S \setminus \{t\}$
    \For{$(t, \psi, t') \in J$ where $t' \in S'$}
    \State $C \gets \Call{Attrs}{\psi}$
    \State $L \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[t] \}$
    \State $R \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[S'] \}$
    \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$
    \EndFor
    \EndFor
    \EndFor
    \EndFor
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
%\end{algorithm}

>>>>

%\begin{algorithm}
  \begin{algorithmic}
    \Require{$q$ is a query plan, $C_l$ is a set of attributes that must be imputed in the output of this query plan, $C_g$ is the set of attributes which are used in the final plan.}
    \Ensure{A set of query plans $Q$ with added imputation.}
    \Function{AddImpute}{$q, C_l, C_g$}
    \State $D_{must} \gets \Call{Dirty}{q} \cap C_l$
    \State $D_{may} \gets D_{must} \cup (\Call{Dirty}{q} \cap C_g)$
    \State $Q \gets \emptyset$
    \If{$D_{must} = \emptyset$}
    \State $Q \gets Q \cup \{q\}$
    \Else
    \State $Q \gets Q \cup \{\mu_{D_{must}}(q),\ \delta_{D_{must}}(q)\}$
    \EndIf

@@

    \Require{$t$ is a table, $\phi$ is a filter predicate, $C_g$ is a set of attributes which are used in the final plan.}
    \Ensure{A set of plans for filtering $t$ with $\phi$.}
    \Function{OptFilter}{$t, \phi, C_g$}
    \State \Return $\{\sigma_{\phi}(q)\ |\ q \in \Call{AddImpute}{t, \Call{Attrs}{\phi}, C_g}\}$
    \EndFunction

@@

    \Require{$q$ is a query plan, $G$ is a set of grouping attributes, $A$ is an aggregation function, $C_g$ is a set of attributes which are used in the final plan.}
    \Ensure{A set of plans for grouping \& aggregating with $G$ and $A$.}
    \Function{OptGroupBy}{$q, G, A, C_g$}
    \State $C \gets G \cup \Call{Attrs}{A}$
    \State $Q' \gets \Call{AddImpute}{q, C, C_g}$
    \State \Return $\{\Call{GroupBy}{q', G, A} ~|~ q' \in Q'\}$
    \EndFunction

@@

    \Require{A set of tables $T$, a plan cache $Q$ containing at least one query plan for each table in $T$, and a relation $J : T \times \Psi \times T$ between tables and join predicates, $C_g$ is the set of attributes which are used in the final plan.}
    \Ensure{An updated plan cache $Q'$ which contains at least one plan that combines all of the tables in $T$.}
    \Function{OptJoin}{$Q, T, J, C_g$}
    \For{$size \in 2...|T|$}
    \For{$S \in \{\text{all length}\ size\ \text{subsets of}\ T\}$}
    \For{$t \in S$}
    \State $S' \gets S \setminus \{t\}$
    \For{$(t, \psi, t') \in J$ where $t' \in S'$}
    \State $C \gets \Call{Attrs}{\psi}$
    \State $L \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[S'] \}$
    \State $R \gets \{\Call{AddImpute}{q, C, C_g} ~|~ q \in Q[t] \}$
    \State $Q[S] \lhd \{l \bowtie_\psi r ~|~ l \in L, r \in R\}$
    \EndFor
    \EndFor
    \EndFor
    \EndFor
    \EndFunction
  \end{algorithmic}
  \caption{Base algorithms for query planning with imputations.}
  \label{algo:plan-helpers}
%\end{algorithm}
